\documentclass[11pt,twoside,openright,letterpaper]{memoir}
\setsecnumdepth{subsection} % include numbering on subsections
\setcounter{tocdepth}{2}

% Use 1/2-inch margins.
% \usepackage[margin=1in,lmargin=1.5in,showframe]{geometry} % ,showframe % for margin boundary box
\usepackage[margin=1in,lmargin=1.5in]{geometry} % ,showframe % for margin boundary box

\usepackage{titling}
\usepackage{helvet} % Use Helvetica font.
\renewcommand{\familydefault}{\sfdefault}
% \usepackage{setspace}
\usepackage{tabularx}
\usepackage{array} % get left justified columns
\newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}} % get left justified columns
\usepackage{pdfpages}
\raggedbottom % solve inconsistent paragraph spacing issue

%%% bibliography:
\usepackage[super,comma,sort&compress,sectionbib]{natbib}
\usepackage{chapterbib}
% switch bibliography numbering to 1. from [1] :
\makeatletter \renewcommand\@biblabel[1]{#1.} \makeatother
\setlength{\bibsep}{0pt}

% math:
\usepackage{amsmath} % for multline
\usepackage{mathtools}
\usepackage{siunitx} % for math formatting
\usepackage{newtxtext,newtxmath} % formats numbers in math mode to consistent font
\newcases{cdcases}{\quad}{$\hfil\displaystyle{##}$\hfil}{$\displaystyle{##}$\hfil}{\lbrace}{.} % for gene conv model

% graphics:
\usepackage{graphicx}
\usepackage{epstopdf} %converting to PDF
\usepackage[section]{placeins} % keep floats within sections
\usepackage{afterpage} % wrap around floats to have them on their own pages
\usepackage{wrapfig} % allow wrapfig and wraptable

% glossary:
\usepackage[toc,nonumberlist]{glossaries}
\makeglossaries

\usepackage{hyperref} % allow urls
\hypersetup{
    pdfauthor={CLCampbell},
%    pdfsubject={Some subject},
    pdftitle={},
%    pdfkeywords={LaTeX, PDF, hyperlinks}.
    citebordercolor = {1 1 1}, % no box around citations
    %,pdfborder = {0 0 1} % visible box around hyperlinks.
    %,pdfborderstyle={/S/U/W 1}
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black, % blue
    pdfpagemode=UseOutlines
}
% \usepackage{hypcap} % anchor links at image, not caption
\usepackage{bookmark} % fix pdf bookmark generation

\title{Patterns of crossing over and gene conversion in meiotic recombination}

\author{Christopher L Campbell}
%Albert Einstein College of Medicine}

\makeatletter\let\Title\@title\makeatother
\makeatletter\let\Author\@author\makeatother

% \newenvironment{abstract}%
% {\cleardoublepage\null \vfill\begin{center}%
% \bfseries \abstractname \end{center}}%
%     {\vfill\null}

\usepackage{tocbibind} % include toc within toc
\usepackage{titlesec}

\newcommand*\captionTitle[2]{\caption[#1]{#1#2}} % figure caption includes short title

\newenvironment{titemize} {
	\begin{itemize}
	  \setlength{\itemsep}{0cm}%
	  \setlength{\parskip}{0cm}%
} {\end{itemize}}

\newenvironment{tenumerate} {
	\begin{enumerate}
	  \setlength{\itemsep}{0cm}%
	  \setlength{\parskip}{0cm}%
} {\end{enumerate}}

\newcommand{\beginsupplement}{ % enable supplement numbering within chapters
    \setcounter{table}{0}
    \renewcommand{\thetable}{\thechapter.S\arabic{table}}%
    \setcounter{figure}{0}
    \renewcommand{\thefigure}{\thechapter.S\arabic{figure}}%
}
\newcommand{\beginmain}{ % reset supplement numbering within chapters
    \setcounter{table}{0}
    \renewcommand{\thetable}{\thechapter.\arabic{table}}%
    \setcounter{figure}{0}
    \renewcommand{\thefigure}{\thechapter.\arabic{figure}}%
}
\newcommand{\beginappendixA}{ % enable appendix numbering of figures
    \setcounter{table}{0}
    \renewcommand{\thetable}{A.\arabic{table}}%
    \setcounter{figure}{0}
    \renewcommand{\thefigure}{A.\arabic{figure}}%
}
%%% adjust spacing between figure number and label in listoffigures:
\makeatletter \renewcommand*\l@figure{\@dottedtocline{1}{1em}{3.2em}} \makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \setcounter{tocdepth}{5}
% \setcounter{page}{1}

\frontmatter

%\begin{titlepage}
\begin{titlingpage}
\begin{center}
    \vspace*{1cm}
    \Large{ \textbf{\Title} } \\
    \vspace{1.5cm}
    \normalsize
    \Author \\ \vspace{2cm}
    \begin{tabularx}{\textwidth}{@{}X@{}X@{}}
        \begin{minipage}[t]{\linewidth}
            \textbf{Candidate:} \\ 
            %\vspace{1.5cm} \hrule\smallskip Signature
            \vspace{1.5cm} \hrule width 0.9\textwidth \smallskip Signature
            %\vspace{1.5cm} \noindent\rule{0.9\textwidth}{0.4pt}
        \end{minipage}
        & \\
        \begin{minipage}[t]{\linewidth}
            \vspace{1cm}
            \textbf{Thesis Advisor:} \\ 
            \vspace{1.5cm} \hrule width 0.9\textwidth \smallskip Signature
            \begin{flushleft}
                Adam Auton, D.Phil. \\
                \small \smallskip
                Assistant Professor, Department of Genetics \\
                Assistant Professor, Department of Epidemiology \& Population Health \\
            \end{flushleft}
        \end{minipage}
        \vspace{1cm}
        &
        \begin{minipage}[t]{\linewidth}
            \vspace{1cm}
            \textbf{Co-advisor:} \\ 
            \vspace{1.5cm} \hrule width 0.9\textwidth \smallskip Signature
            \begin{flushleft}
                Bernice E. Morrow, Ph.D. \\
                \small \smallskip
                Professor, Department of Genetics \\
                Professor, Department of Obstetrics \& Gynecology and Women's Health \\
                Professor, Department of Pediatrics (Cardiology) \\
                Sidney L. and Miriam K. Olson Chair in Cardiology \\
                Director, Division of Translational Genetics, Department of Genetics \\
            \end{flushleft}
        \end{minipage}\\
    \end{tabularx}
%%%
\vfill
Submitted in partial fulfillment of the requirements for the Degree of Doctor of
Philosophy in the Graduate Division of Medical Sciences.
\vspace{0.5cm}
\\ Albert Einstein College of Medicine
\\ Yeshiva University
\\ New York
\\ May 31, 2016
\end{center}
%\end{titlepage}
\end{titlingpage}
%\newpage
\setcounter{page}{1}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\chapter{Abstract}
\addcontentsline{toc}{chapter}{Abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abstract: The abstract of the Dissertation is to include: a hypothesis, the procedures followed, the
% significant results and the general conclusions. The abstract is to be presented on a separate page
% headed with the word ABSTRACT in capital letters centered on the page. On the next line is the
% title of the Dissertation. The following line is the full name of the student. The length of the
% abstract must not exceed 600 words. (Please note the separate instructions for the 350 word
% microfilm copy abstract described in the first section of this manual.)
\thispagestyle{plain}
\begin{centering} ABSTRACT \\ \vspace{10pt} %\end{centering}
\Title \\
\Author \\
\end{centering}
\vspace{10pt}
 \DoubleSpacing


Recombination during meiosis is an essential process to the generation of gametes.
It acts to shuffle genetic variation within the genome, generating new combination of alleles that can be subject to natural selection.
Failures of the chromosomes to synapse and recombine can result in aneuploidy, while improper repair of double strand breaks can cause chromosome rearrangements, both of which are highly deleterious.

In humans and other mammals, recombination has been studied using a wide variety of approaches, including molecular methods such as sperm typing, and inferential methods, such as linkage disequilibrium modeling and pedigree analysis.
Recombination has two outcomes: crossover and gene conversion.
In crossover, there is an equal exchange of genetic material between homologous chromosomes, while gene conversion is the non-reciprocal transfer of genetic information between homologues that is limited to smaller intervals.
Most research has focused on crossover, since identifying gene conversion events can be methodologically challenging.
Crossovers vary widely in frequency and placement within the genome as well as between individuals, sexes, populations, and species.
Multiple factors affect the placement of crossovers within the genome, including the PRDM9 protein, which has been shown to localize crossovers to concentrated areas of recombination known as hotspots.
In addition, crossover interference acts to inhibit crossovers in close proximity.
It has been suggested that these recombination properties change with age, which might lead to chromosomal aneuploidy.

% chapter 2
% Despite the importance of recombination, it is a relatively rare process, with only a few dozen recombination events expected to occur within a given meiosis. In order to better understand recombination in humans I have employed statistical methods to identify sex specific differences in recombination.
In Chapter \ref{ch:cointEsc}, I performed a pedigree analysis using data derived from over 18,000 meioses.
Using this data, I characterized sex differences in human recombination, and created sex-specific maps of recombination in the human genome.
These results extend previous work showing that females have a higher recombination rate than males, by a ratio of 1.6 in the autosomes, and that male recombination tends to be focused towards the telomeres.
In addition, I find that males and females have a difference in the proportion of crossovers that overlap known hotspots, with males having 4.6\% higher hotspot overlap.
Finally, I measure crossover interference, which affects the spatial positioning of crossovers.
I find that in older mothers there is a steep increase in crossovers that escape regulation by interference, and appear closely spaced, pointing to a possible deregulation of recombination that increases with age in females.


In Chapter \ref{ch:cointExtras}, I present unpublished data, an extension to Chapter \ref{ch:cointEsc}, which confirms the age effects using samples from older individuals.
Additionally, I include a re-analysis of public data from single cell sperm and oocytes, showing the interference varies widely on an individual basis.

In Chapter \ref{ch:dogPed}, I focused on crossover patterns using a complex pedigree of inbred domestic dogs, consisting of 408 meioses.
Dogs are unique in that their PRDM9 orthlog has accumulated mutations, rendering it inactive, and raising question as to how crossovers are placed without this regulatory mechanism.
I found that female dogs have a higher recombination rate than males but to a lesser extent than humans.
% , a ratio of 1.2 within the autosomes, and that 
As in humans, male recombination in dogs strongly prefers the telomeric regions.
Despite the absence of PRDM9, dog recombination appears to be more concentrated to a smaller proportion of sequence when compared to humans, suggesting the existence of hotspots.
Finally, I show evidence for positive crossover interference acting in the dog genome with a similar mechanism to what has been seen in humans.

In Chapter \ref{ch:geneConv}, I examined gene conversion, using a new method that I developed.
In this chapter, I developed a novel computational approach using a hidden Markov model to detect gene conversion events in admixed population genetic data.
The model uses two divergent ancestral reference populations in order to predict the location of gene conversion events on an admixed haplotype.
Simulation results show that the model can plausibly be used to detect gene conversion.
% , however the complexity is high, with a large number of states at each site.

Overall, I have performed three separate but interrelated studies of recombination that expand on existing studies.
First, my work expands on the current knowledge of recombination variation and provides evidence for sexual dimorphism in crossover placement and frequency.
Second, my work has further investigated the effects of the loss of PRDM9 on the recombination landscape in dogs, and through comparison to humans, provided insight into recombination in both species.
Finally, I developed a novel computational approach for investigating gene conversion, in order to enable further understanding of the properties of genetic recombination.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Acknowledgements}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The Training Program in Cellular and Molecular Biology and Genetics, T32 GM007491.

\SingleSpacing
\clearpage
\tableofcontents
\listoffigures
\listoftables

\glsaddall
\clearpage\begingroup\let\newpage\relax
    \printglossary[type=main,style=long,nonumberlist,title={List of Abbreviations}]
    \addcontentsline{toc}{chapter}{List of Abbreviations}
\endgroup
\include{abbrv}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\mainmatter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapterstyle{ger}
 \DoubleSpacing
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Introduction} \label{ch:introduction}
\include{introduction/introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{SingleSpace}
% \chapter[Title displayed in ToC][Title displayed in header]{Title}
\chapter[Escape from crossover interference increases with maternal age][Interference escape increases with maternal age]{Escape from crossover interference increases with maternal age} \label{ch:cointEsc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\noindent Christopher L. Campbell$^1$, Nicholas A. Furlotte$^2$, Nick Eriksson$^{2,\dagger}$, David Hinds$^2$, and Adam Auton$^1$

\vspace{0.5cm}
\noindent This manuscript has been published: \\
Campbell, C. L. et al. Escape from crossover interference increases with maternal age. Nat. Commun. 6:6260 doi: 10.1038/ncomms7260 (2015). \\
PMCID: PMC4335350

\vspace{0.5cm}
\noindent $^1$ Department of Genetics, Albert Einstein College of Medicine, 1301 Morris Park Avenue, Bronx, New York 10461, USA. \\
\noindent $^2$ 23andMe Inc., Mountain View, California 94043, USA. \\
\noindent $^\dagger$ Present address: Coursera, 381 East Evelyn Avenue, Mountain View, California 94041, USA. \\

\vspace{0.5cm}
\begin{centering}
    Correspondence and requests for materials should be addressed to \\
    A.A.\ (email: \texttt{adam.auton@einstein.yu.edu}). \\
\end{centering}
\end{SingleSpace}

\include{cointEscape/cointEscape}
\include{cointEscape/cointEscapeSupp}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{SingleSpace}
\chapter{Crossover interference varies by age and individual} \label{ch:cointExtras}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\noindent Christopher L. Campbell$^1$ and Adam Auton$^{1*}$

\vspace{0.5cm}
\noindent This chapter contains unpublished data.\\

\vspace{0.5cm}
\noindent $^1$ Department of Genetics, Albert Einstein College of Medicine, 1301 Morris Park Avenue, Bronx, New York 10461, USA. \\
\noindent $^*$ Former affiliation.
\end{SingleSpace}

\include{cointExtras/cointExtras}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{SingleSpace}
\chapter[A pedigree-based map of recombination in the domestic dog genome][Recombination maps in the dog genome]{A pedigree-based map of recombination in the domestic dog genome} \label{ch:dogPed}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\noindent Christopher L Campbell$^1$, Claude Bh\'{e}rer$^{1*,2}$, Bernice E Morrow$^1$, Adam R Boyko$^3$, and Adam Auton$^{1*}$

\vspace{0.5cm}
\noindent This manuscript has been submitted to \textit{Genetics}. \\

\vspace{0.5cm}
\noindent $^1$ Department of Genetics, Albert Einstein College of Medicine, 1301 Morris Park Avenue, Bronx, New York 10461, USA. \\
\noindent $^2$ New York Genome Center, New York, New York 10013, USA \\
\noindent $^3$ Department of Biomedical Sciences, College of Veterinary Medicine, Cornell University, Ithaca, New York 14853, USA \\
\noindent $^*$ Former affiliation.

\vspace{0.5cm}
\begin{centering}
    Correspondence and requests for materials should be addressed to \\
    A.A.\ (email: \texttt{adam.auton@gmail.com}). \\
\end{centering}
\end{SingleSpace}

\include{dogPed/dogPed}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{SingleSpace}
\chapter[Detection of gene conversion in human admixed population genetic data][A model for the detection of gene conversion]{Detection of gene conversion in human admixed population genetic data} \label{ch:geneConv}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\noindent Christopher L. Campbell$^1$ and Adam Auton$^{1*}$

\vspace{0.5cm}
\noindent This chapter contains unpublished data.\\

\vspace{0.5cm}
\noindent $^1$ Department of Genetics, Albert Einstein College of Medicine, 1301 Morris Park Avenue, Bronx, New York 10461, USA. \\
\noindent $^*$ Former affiliation.
\end{SingleSpace}

\include{geneConv/geneConv}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Discussion} \label{ch:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\include{discussion/discussion}

\backmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{SingleSpace}
\renewcommand{\thechapter}{A}
\chapter{Appendix A} \label{ch:appendixCB}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\vspace{0.5cm}
\noindent Results presented in this appendix are from a manuscript under review at \textit{Nature Communications}:

\vspace{0.5cm}
\noindent Refined genetic maps reveal sexual dimorphism in human meiotic recombination at multiple scales. \\

\noindent Claude Bh\'{e}rer$^{1*,2}$, Christopher L Campbell$^1$, and Adam Auton$^{1*}$.

\vspace{0.5cm}
\noindent $^1$ Department of Genetics, Albert Einstein College of Medicine, 1301 Morris Park Avenue, Bronx, New York 10461, USA. \\
\noindent $^2$ New York Genome Center, New York, New York 10013, USA \\
\noindent $^3$ Department of Biomedical Sciences, College of Veterinary Medicine, Cornell University, Ithaca, New York 14853, USA \\
\noindent $^*$ Former affiliation. \\
\noindent $^\#$ Present address.

\vspace{0.5cm}
\begin{centering}
    Corresponding authors: \\
    Adam Auton (\texttt{adam.auton@gmail.com}) \\
    Claude Bh\'{e}rer (\texttt{claudebherer@gmail.com}) \\
\end{centering}
\end{SingleSpace}

\include{appendix_cb/appendix}
%\includepdf[pages=1-20]{appendix_papers/SexDimorphismRecombination_2016April8_maintext.pdf}
%\includepdf[pages=1-21]{appendix_papers/SexDimorphismRecombination_2016April8_SUPP.pdf}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{Appendix B}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\includepdf[pages=1-7]{appendix_papers/Campbell2015.pdf}
%\includepdf[pages=1-28]{appendix_papers/Campbell2015_supp.pdf}
%\includepdf[pages=1-20]{appendix_papers/nature15393.pdf}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
